Discovering constraints in spreadsheets and tabular data

### 2016-09-08 Paper/ECAI/tabular.tex
Expanded runtime analysis

For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of size~$k$, the worst case complexity becomes $O(n * m * k)$.
However, in practice, most subassignments will not satisfy the constraint template.
Since our implementation tries to account for this by failing fast, the amortized time is constant in the size of the vectors~$k$, unless $k$~is very large or many constraints hold in the data.
For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small).

The benchmark experiments show that the number of cells in a spreadsheet has a very limited impact on the running time.
Synthetic experiments with random integers and strings show that increasing the size of the vectors has a neglegible impact on the performance.

>>>>

For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of size~$k$, the worst case complexity becomes $O(n * m * k)$.
However, in practice, most subassignments will not satisfy the constraint template.
Since our implementation tries to account for this by failing fast, the amortized time is constant in the size of the vectors~$k$, unless $k$~is very large\footnote{Specifically, if $k$ is much larger than the number of candidate subassignments} or many constraints hold in the data.
For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small).

The benchmark experiments show that the number of cells in a spreadsheet has a very limited impact on the running time.
Synthetic experiments with random integers and strings show that increasing the size of the vectors has a negligible impact on the performance for most templates.

Aggregate constraints, e.g. sum or max, have an argument that takes subblocks of varying sizes.
Therefore, the size of the blocks in the spreadsheet influences the runtime of these constraints.
For a block of size~$m$ there are $\frac{m(m + 1)}{2} = O(m^2)$ subblocks, thus, if a spreadsheet has smaller sized blocks, the search for aggregate constraints will be faster.

Synthetic experiments with random integers confirm that splitting a constant number of vectors into smaller blocks improves the performance of the aggregate constraint search.

These experiments, however, also show that the running time for various non-aggregate constraints increases as the size of blocks becomes very small.
Since these constraints have single vector arguments the block size should not have a strong effect.
However, in practice, for very small blocks to the superblock assignment phase has to generate a very large number of combinations, while for larger blocks this work is done mostly during the subassignment generation phase.
Since this last phase is more adept at generating these combinations, shifting this workload to the assignment phase slows down the algorithm.
